---
title: "Backward Compatibility: Working with Legacy Code"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Backward Compatibility: Working with Legacy Code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

if (!"MVBeliefUpdatr" %in% loadedNamespaces()) {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    tryCatch(pkgload::load_all("..", quiet = TRUE), error = function(e) NULL)
  }
}
```

# Introduction and Overview

Many existing pipelines and user scripts built against pre-v1.0 versions of **MVBeliefUpdatr** expect models and templates to be formatted as `tibble` data frames with list-columns of parameters (`mu`, `Sigma`, `m`, `S`, etc.).

To provide a non-breaking migration path, **MVBeliefUpdatr** implements a bidirectional compatibility bridge in `R/S7-to-legacy-tibble-compatibility.R`.

```{r setup}
library(MVBeliefUpdatr)
library(tibble)
library(dplyr)
```

---

# 1. Coercing S7 Objects with `as_tibble()` {#as-tibble}

Calling `as_tibble()` on any S7 representation, template, or cognitive model produces a tibble matching the pre-S7 column schema, names, types, and column order:

```{r as_tibble_bridge}
# Create an S7 cognitive model
observer_model <- example_mvg_ideal_observer()

# Coerce an S7 cognitive model to a legacy tibble
model_tibble <- as_tibble(observer_model)
model_tibble

# Inspect column names and list-column contents
names(model_tibble)
model_tibble$mu
model_tibble$Sigma
```

When called, `as_tibble()` issues an informative deprecation warning noting that S7 objects are the canonical format and that some metadata (such as noise behavior and internal validators) are lost in tabular representations.

---

# 2. Deprecated `dplyr` Verb Forwarding {#dplyr-forwarding}

For workflows that previously called `dplyr` verbs directly on model objects (e.g., `model %>% mutate(...)` or `model %>% filter(...)`), S3 method forwarding intercepts these calls, converts the object to a tibble, applies the requested verb, and informs the user:

```{r dplyr_forwarding}
# Legacy workflow: calling filter() or select() on an S7 model
filtered_df <- suppressWarnings(filter(observer_model, category == "/i/"))
filtered_df

selected_df <- suppressWarnings(select(observer_model, category, mu, Sigma))
selected_df
```

---

# 3. API Migration Mapping Table {#migration-mapping-table}

The table below maps legacy (pre-v0.1.0) functions to their modern v0.1.0+ S7 API replacements. All legacy functions issue deprecation warnings in v0.1.0 and are scheduled for removal in v0.2.0.

| Legacy API Function (pre-v0.1.0) | Modern S7 API Replacement (v0.1.0+) | Description / Category |
| :--- | :--- | :--- |
| `make_MVG_from_data()` | `new_mvg_category_representation_template_from_data()` | Estimate MVG category template from data |
| `make_MVG_ideal_observer_from_data()` | `new_mvg_ideal_observer_from_data()` | Construct MVG cognitive model directly from data |
| `make_NIW_belief_from_data()` | `new_niw_category_representation_template_from_data()` | Estimate NIW belief template from data |
| `make_NIW_ideal_adaptor_from_data()` | `new_niw_ideal_adaptor_from_data()` | Construct NIW ideal adaptor model from data |
| `make_exemplars_from_data()` | `new_exemplar_category_representation_template_from_data()` | Estimate Exemplar category template from data |
| `make_exemplar_model_from_data()` | `new_exemplar_model_from_data()` | Construct Exemplar model from data |
| `lift_MVG_to_MVG_ideal_observer()` | `lift_representation_to_model()` / `new_mvg_ideal_observer()` | Wrap MVG category template in cognitive model |
| `lift_NIW_belief_to_NIW_ideal_adaptor()` | `lift_representation_to_model()` / `new_niw_ideal_adaptor()` | Wrap NIW template in ideal adaptor model |
| `lift_exemplars_to_exemplar_model()` | `lift_representation_to_model()` / `new_exemplar_model()` | Wrap Exemplar template in cognitive model |
| `is.ideal_adaptor_stanfit()` | `S7::S7_inherits(x, IdealAdaptorStanfit)` | Type check for S7 Stanfit objects |
| `read_ideal_adaptor_stanfit()` | `read_stanfit()` | Read saved Stanfit object from RDS |
| `write_ideal_adaptor_stanfit()` | `write_stanfit()` | Write Stanfit object to RDS |
| `get_info_from_model()` | `get_parameters()`, `summary()`, or `as_tibble()` | Extract model parameters / metadata |
| `sample_observation()` | `sample_observations()` | Draw random samples from category representations |

---

# 4. Migration Recommendations {#migration-recommendations}

When modernizing legacy code to the S7 architecture:

1. **Replace `as_tibble()` calls with direct S7 methods:** Instead of extracting parameters by indexing list-columns (`model_df$mu[[1]]`), use `get_parameters(model, statistic = "mu")` or `get_expected_mu(model)`.

2. **Use S7 constructors directly:** Replace legacy `make_*` and `lift_*` functions with `new_*_category_representation()` and `new_*_ideal_observer()`.

3. **Dispatch evaluation directly:** Call `categorize(model, data)` and `posterior(model, data)` directly on S7 objects rather than passing tibbles to legacy wrappers.

