---
title: "Fitting and Working with MVBeliefUpdatr Stanfit Models"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Fitting and Working with MVBeliefUpdatr Stanfit Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

if (!"MVBeliefUpdatr" %in% loadedNamespaces()) {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    tryCatch(pkgload::load_all("..", quiet = TRUE), error = function(e) NULL)
  }
}
```

<script>
document.addEventListener('DOMContentLoaded', function() {
  var tocLinks = document.querySelectorAll('#TOC a[href^="#"]');
  tocLinks.forEach(function(anchor) {
    anchor.addEventListener('click', function(e) {
      var href = this.getAttribute('href');
      var targetId = href.substring(1);
      var targetEl = document.getElementById(targetId) || 
                     document.getElementById('id_' + targetId) || 
                     document.querySelector('[name="' + targetId + '"]');
      if (targetEl) {
        e.preventDefault();
        targetEl.scrollIntoView({ behavior: 'smooth', block: 'start' });
        if (history.pushState) {
          history.pushState(null, null, href);
        }
      }
    });
  });
});
</script>

# 1. Introduction and Overview {#introduction-and-overview}

The **MVBeliefUpdatr** package provides Bayesian hierarchical Stan models for incremental cognitive belief-updating during speech perception and multidimensional categorization experiments. These models fit observer and adaptor beliefs (e.g., Normal-Inverse-Wishart [NIW] and Normal-Inverse-$\chi^2$ [NIX]) across exposure blocks or participant groups using MCMC sampling via **rstan**.

Fitted models are returned as typed S7 objects inheriting from `MVBU_Stanfit` (e.g., `NIW_IdealAdaptorStanfit` or `NIX_IdealAdaptorStanfit`). This vignette details:

1. **Preparing Data and Inputs**: Structuring exposure data and bundling priors with `new_ideal_adaptor_stanfit_input()`.
2. **Model Fitting**: Executing Bayesian estimation via `fit_ideal_adaptor()` or `fit_ideal_observer()`.
3. **Model Summaries & Diagnostics**: Printing, summarizing, and evaluating convergence (`rhat`, `neff_ratio`, `get_draws`).
4. **Accelerated Post-Processing**: Pre-extracting parameter draws with `add_parameter_draws()` for fast visualization.
5. **Model Comparison & Criteria**: Storing information criteria with `add_criterion()`, comparing models with `loo_compare()`, and generating posterior predictive checks with `pp_check()`.

### Companion Vignettes
- [S7 Class Architecture, Constructors, and Workflows](s7-class-structure-and-workflows.html)
- [Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html)
- [Backward Compatibility: Working with Old Code](backward-compatibility-working-with-old-code.html)

```{r setup}
library(MVBeliefUpdatr)
library(tibble)
library(dplyr)
```

---

# 2. Preparing Data and Input Bundles (`MVBU_Staninput`) {#preparing-data-and-inputs}

Before fitting a model in Stan, raw participant observations (containing continuous cue dimensions such as Formant frequencies $F_1, F_2$ or Voice Onset Time [VOT]) are packaged alongside prior hyperparameters using `new_ideal_adaptor_stanfit_input()`.

The input pipeline:
1. Standardizes or scales continuous cue dimensions (e.g., z-scoring relative to exposure distribution).
2. Sets up conjugate prior distributions over category parameters (e.g., mean $m_0$, scatter matrix $S_0$, prior sample size $\kappa_0$, and degrees of freedom $\nu_0$).
3. Creates a strongly typed `IdealAdaptorStanfitInput` or `IdealObserverStanfitInput` S7 object.

```{r prepare_staninput_example, eval = FALSE}
# Create simulated 2-cue exposure data for two vowel categories (/i/ and /e/)
exposure_data <- tibble::tibble(
  F1 = c(rnorm(20, mean = 300, sd = 30), rnorm(20, mean = 500, sd = 40)),
  F2 = c(rnorm(20, mean = 2200, sd = 100), rnorm(20, mean = 1800, sd = 120)),
  category = factor(rep(c("cat_A", "cat_B"), each = 20)),
  group = "exposure",
  subject = 1L
)

# Bundle data and priors into an S7 Staninput object
stan_input <- new_ideal_adaptor_stanfit_input(
  exposure = exposure_data,
  test = exposure_data,
  cues = c("F1", "F2"),
  category = "category",
  stanmodel = "NIW_ideal_adaptor"
)
```

---

# 3. Fitting Bayesian Models (`fit_ideal_adaptor`) {#fitting-bayesian-models}

To fit the model, `fit_ideal_adaptor()` passes the `stan_input` object to Stan's MCMC sampler via **rstan**.

```r
# Executing MCMC sampling (interactive fitting call)
fit <- fit_ideal_adaptor(
  stan_input,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 42
)
```

For reproducible vignette compilation without long MCMC sampling delays, pre-compiled model fits can be accessed via package fixtures or helpers:

```{r load_example_fit}
# Load pre-computed 2-cue NIW fit fixture
fit_file <- system.file("extdata", "example-stanfit-NIW_ideal_adaptor-1-standardize-42.rds", package = "MVBeliefUpdatr")
if (!file.exists(fit_file)) {
  fit_file <- file.path("..", "tests", "testthat", "models", "example-stanfit-NIW_ideal_adaptor-1-standardize-42.rds")
}
fit <- readRDS(fit_file)

# Inspect class hierarchy
class(fit)
get_model_type(fit)
```

---

# 4. Printing, Summarizing, and Diagnostics {#printing-summarizing-diagnostics}

`MVBU_Stanfit` objects provide clear, structured S7 `print()` and `summary()` outputs.

## Printing
Printing an `MVBU_Stanfit` object displays its S7 class, model type, cue labels, category labels, transformation status, and MCMC sampling specifications:

```{r print_fit}
print(fit)
```

## Summarizing
`summary()` computes posterior summary statistics for prior hyperparameters ($\kappa, \nu, m, S$) and category posterior distributions across groups:

```{r summary_fit}
summary(fit)
```

## Extracting Draws & MCMC Diagnostics
Posterior draws can be extracted in tidy format using `get_draws()`. Standard MCMC convergence diagnostics ($\widehat{R}$ and effective sample size ratio $n_{eff}/N$) are accessible via `rhat()` and `neff_ratio()`:

```{r diagnostics}
# Tidy posterior draws
draws_df <- get_draws(fit, summarize = FALSE)
head(draws_df)

# Convergence diagnostics across parameters
head(rhat(fit))
head(neff_ratio(fit))
```

---

# 5. Pre-extracting Parameter Draws (`add_parameter_draws`) {#pre-extracting-draws}

When producing complex multi-panel category plots, categorization decision boundaries, or posterior density curves across multiple exposure trials, downstream plotting routines frequently need to access parameter draw matrices ($\mu, \Sigma, m, S$).

By default, extracting draws repeatedly from an `MVBU_Stanfit` object requires untransforming cue coordinates and pivoting wide tables into matrix formats on every call.

Calling **`add_parameter_draws()`** pre-extracts and indexes parameter matrices once and stores them in the object's `@cache` slot using functional copy-on-modify semantics:

```{r add_parameter_draws_demo}
# Pre-extract draw matrices into the model's @cache slot
fit <- add_parameter_draws(fit)

# Inspect cached draw matrices
cache_contents <- .get_cache(fit)$draw_matrices
names(cache_contents)
cache_contents$n_draws
```

### Why Precomputed Parameters are Advantageous
1. **Performance**: Eliminates repeated data-wrangling loops during high-resolution contour or grid evaluations.
2. **Seamless Visualization**: Plotting functions such as `plot_categories()` and `plot_categorization_by_cue()` automatically leverage cached draw matrices when present.

For comprehensive examples of plotting category distributions and decision boundaries, see the companion vignette:
[Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html).

---

# 6. Model Comparison and Information Criteria {#model-comparison-and-criteria}

`MVBU_Stanfit` objects integrate with **loo** and **bayesplot** for Bayesian model comparison, cross-validation, and posterior predictive checking, following the argument conventions of **brms**.

## Adding Fit Criteria (`add_criterion`)

Model fit criteria (such as `"loo"`, `"waic"`, `"bayes_R2"`, `"loo_R2"`, `"kfold"`, `"loo_subsample"`, or `"marglik"`) are computed and stored in the model's `@criteria` slot using `add_criterion()`:

```{r add_criterion_demo}
# Compute and store LOO and WAIC criteria
fit <- add_criterion(fit, criterion = c("loo", "waic", "bayes_R2"), model_name = "NIW_2Cue_Model")

# Inspect stored criteria
names(fit@criteria)
fit@criteria$loo
fit@criteria$bayes_R2
```

## Model Comparison (`loo_compare`)

When evaluating alternative model specifications (e.g., comparing different prior specifications or model formulations), `loo_compare()` performs Leave-One-Out Cross-Validation model comparison via `loo::loo_compare()`:

```{r loo_compare_demo}
# Duplicate model for demonstration of loo_compare signature
fit2 <- fit
fit2@metadata$model_name <- "NIW_Alternative_Model"

# Compare models using LOO-CV
comp <- loo_compare(fit, fit2, criterion = "loo")
comp
```

> **Tip:** Leave-One-Out Cross-Validation (`loo`) is recommended over `waic` for robust model comparison, as `waic` can be sensitive to influential observations.

## Posterior Predictive Checking (`pp_check`)

Graphical posterior predictive checks compare observed exposure data $y$ with synthetic observations $y_{rep}$ drawn from the model's posterior predictive distribution:

```{r pp_check_demo}
if (requireNamespace("bayesplot", quietly = TRUE)) {
  # Overlay observed response density with posterior predictive replicates
  pp_check(fit, type = "dens_overlay", ndraws = 30)
}
```

---

# 7. Summary and Related Documentation {#summary}

In summary, the `MVBU_Stanfit` workflow provides an integrated pipeline for Bayesian belief updating:

```
new_ideal_adaptor_stanfit_input() ──► fit_ideal_adaptor() ──► print() / summary()
                                                                      │
                                                                      ├──► add_parameter_draws() (Fast Plotting)
                                                                      ├──► add_criterion() & loo_compare() (Model Selection)
                                                                      └──► pp_check() (Predictive Checks)
```

### Links to Related Vignettes
- [S7 Class Architecture, Constructors, and Workflows](s7-class-structure-and-workflows.html): Detailed technical reference for S7 classes, generics, and model construction.
- [Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html): Comprehensive plotting guide for 1D, 2D, and 3D category distributions and decision boundaries.
- [Backward Compatibility: Working with Old Code](backward-compatibility-working-with-old-code.html): Transition guide from legacy list-column models to S7 architecture.
