---
title: "Fitting and Working with MVBeliefUpdatr Stanfit Models"
author: "T. Florian Jaeger"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Fitting and Working with MVBeliefUpdatr Stanfit Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

<script>
document.addEventListener('DOMContentLoaded', function() {
  var tocLinks = document.querySelectorAll('#TOC a[href^="#"]');
  tocLinks.forEach(function(anchor) {
    anchor.addEventListener('click', function(e) {
      var href = this.getAttribute('href');
      var targetId = href.substring(1);
      var targetEl = document.getElementById(targetId) || 
                     document.getElementById('id_' + targetId) || 
                     document.querySelector('[name="' + targetId + '"]');
      if (targetEl) {
        e.preventDefault();
        targetEl.scrollIntoView({ behavior: 'smooth', block: 'start' });
        if (history.pushState) {
          history.pushState(null, null, href);
        }
      }
    });
  });
});
</script>

# 1. Introduction and Overview {#introduction-and-overview}

The **MVBeliefUpdatr** package provides Bayesian hierarchical Stan models for incremental cognitive belief-updating during speech perception and multidimensional categorization experiments. These models fit observer and adaptor beliefs (e.g., Normal-Inverse-Wishart [NIW] and Normal-Inverse-$\chi^2$ [NIX]) across exposure blocks or participant groups using MCMC sampling via **rstan**.

Fitted models are returned as typed S7 objects inheriting from `MVBU_Stanfit` (e.g., `NIW_IdealAdaptorStanfit` or `NIX_IdealAdaptorStanfit`). This vignette details:

1. **Preparing Data and Inputs**: Structuring exposure data and bundling priors with `new_ideal_adaptor_stanfit_input()`.
2. **Model Fitting**: Executing Bayesian estimation via `fit_ideal_adaptor()` or `fit_ideal_observer()`.
3. **Model Summaries & Diagnostics**: Printing, summarizing, and evaluating convergence (`rhat`, `neff_ratio`, `get_draws`).
4. **Accelerated Post-Processing**: Pre-extracting posterior latents and closures with `add_posterior_latents()` for fast visualization.
5. **Model Comparison & Criteria**: Storing information criteria with `add_criterion()`, comparing models with `loo_compare()`, and generating posterior predictive checks with `pp_check()`.

### Companion Vignettes
- For the basic class structure of `MVBeliefUpdatr`, see the vignette [S7 Class Architecture, Constructors, and Workflows](s7-class-structure-and-workflows.html)
- For visualization workflows corresponding to these objects, see the vignette:
[Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html).

```{r setup}
library(MVBeliefUpdatr)
library(tibble)
library(dplyr)
library(loo)
```

---

# 2. Preparing Data and Input Bundles (`MVBU_Staninput`) {#preparing-data-and-inputs}

Before fitting a model in Stan, raw participant observations (containing continuous cue dimensions such as Formant frequencies $F_1, F_2$ or Voice Onset Time [VOT]) are packaged alongside prior hyperparameters using `new_ideal_adaptor_stanfit_input()`.

The input pipeline:
1. Standardizes or scales continuous cue dimensions (e.g., z-scoring relative to exposure distribution).
2. Sets up conjugate prior distributions over category parameters (e.g., mean $m_0$, scatter matrix $S_0$, prior sample size $\kappa_0$, and degrees of freedom $\nu_0$).
3. Creates a strongly typed `IdealAdaptorStanfitInput` or `IdealObserverStanfitInput` S7 object.

```{r prepare_staninput_example, eval = FALSE}
# Construct standard 2-cue NIW ideal adaptor Stanfit input
stan_input <- example_ideal_adaptor_stanfit_input(
  model_family = "NIW",
  n_cues = 2L,
  seed = 42L
)
```

---

# 3. Fitting Bayesian Models (`fit_ideal_adaptor`) {#fitting-bayesian-models}

To fit the model, `fit_ideal_adaptor()` passes the `stan_input` object to Stan's MCMC sampler via **rstan**.

```r
# Executing MCMC sampling (interactive fitting call)
fit <- fit_ideal_adaptor(
  stan_input,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 42
)
```

For reproducible vignette compilation without long MCMC sampling delays, pre-compiled model fits can be accessed via package fixtures or helpers:

```{r load_example_fit}
# Load or fit 1-cue NIW ideal adaptor Stanfit model
models_dir <- if (dir.exists("../tests/testthat/models")) {
  "../tests/testthat/models"
} else if (dir.exists("tests/testthat/models")) {
  "tests/testthat/models"
} else {
  tempdir()
}
fit_file <- file.path(
  models_dir,
  "example-stanfit-NIW_ideal_adaptor-1-standardize-42.rds"
)
fit <- example_ideal_adaptor_stanfit(
  model_family = "NIW",
  n_cues = 1L,
  file = fit_file,
  file_refit = "on_change"
)

# Inspect class hierarchy
class(fit)
get_model_type(fit)
```

---

# 4. Lightweight Posterior Representations (`MVBU_StanfitPosterior`) {#lightweight-posterior-representations}

While full `MVBU_Stanfit` objects store raw sampling chains, diagnostic metrics, compiled C++ structures, and source matrices, downstream evaluation and plotting workflows often only require parameter draw distributions.

The `MVBU_StanfitPosterior` S7 class provides a lightweight representation of fitted posterior distributions. It strips away heavy Stan compilation structures while retaining parameter draws and metadata, enabling fast serialization, evaluation, and visualization.

```{r stanfit_posterior_demo}
# Coerce a full Stanfit model to a lightweight posterior representation
post_model <- as_MVBU_stanfit_posterior(fit)
post_model

# Inspect cue labels, categories, response categories, and groups
get_cue_labels(post_model)
get_category_labels(post_model)
get_response_category_labels(post_model)
get_group_labels(post_model, include_prior = TRUE)
get_model_family(post_model)

# Evaluate decision likelihoods directly on the lightweight posterior
eval_res <- evaluate_model(post_model)
head(eval_res)
```

---

# 5. Printing, Summarizing, and Diagnostics {#printing-summarizing-diagnostics}

`MVBU_Stanfit` objects provide clear, structured S7 `print()` and `summary()` outputs.

## Printing
Printing an `MVBU_Stanfit` object displays its S7 class, model type, cue labels, category labels, transformation status, and MCMC sampling specifications:

```{r print_fit}
print(fit)
```

## Summarizing
`summary()` computes posterior summary statistics for prior hyperparameters ($\kappa, \nu, m, S$), group-specific belief parameters, and full distributions across MCMC draws for both **expected category moments** and **marginal category moments**:

 - **Expected moments**: Moments of the category conditional on the parameters of each draw. The expected covariance is $\mathbb{E}[\mathbf{\Sigma}] = \frac{\mathbf{S}}{\nu - D - 1}$.
 - **Marginal moments**: Moments marginalizing over uncertainty in the category mean $\mu$, scaling the covariance by $\frac{\kappa + 1}{\kappa}$: $\mathbf{\Sigma}_{\text{marg}} = \frac{\kappa + 1}{\kappa} \frac{\mathbf{S}}{\nu - D - 1}$.

Both moment tables report the mean, standard deviation, and standard Bayesian credible quantiles (2.5%, 25%, 50%, 75%, 97.5%) across posterior draws for all cues and cue covariances:

```{r summary_fit}
summary(fit)
```

## Extracting Draws & MCMC Diagnostics
Posterior draws can be extracted in tidy format using `get_draws()`. When `nest = TRUE` is specified, `get_draws()` provides draw-level expected covariance matrices (`Sigma_exp`), marginal covariance matrices (`Sigma_marg`), and standard covariance matrices (`Sigma`).

Standard MCMC convergence diagnostics ($\widehat{R}$ and effective sample size ratio $n_{eff}/N$) are accessible via `rhat()` and `neff_ratio()`:

```{r diagnostics}
# Tidy posterior draws with draw-level covariance matrices
draws_df <- get_draws(fit, nest = TRUE, summarize = FALSE)
names(draws_df)

# Convergence diagnostics across parameters
head(rhat(fit))
head(neff_ratio(fit))
```

---

# 5. Pre-extracting Posterior Latents & Evaluation Closures (`add_posterior_latents`) {#pre-extracting-latents}

When producing complex multi-panel category plots, categorization decision boundaries, or posterior density curves across multiple exposure trials, downstream plotting and evaluation routines frequently need to access posterior latent parameter draws ($m, S, \kappa, \nu, \mathbf{\Sigma}_{\text{exp}}, \mathbf{\Sigma}_{\text{marg}}$) and evaluate likelihood or posterior functions across observation grids.

By default, extracting draws repeatedly from an `MVBU_Stanfit` object requires untransforming cue coordinates and pivoting tables on every call.

Calling **`add_posterior_latents()`** extracts draw-level latents once, computes expected and marginal covariance moments, compiles vectorized evaluation closures, and stores them in the object's `@cache` slot:

 - `draws`:
   - `parameters`: Nested tibble containing draw-level representations of $m, S, \kappa, \nu, \mathbf{\Sigma}_{\text{exp}}, \mathbf{\Sigma}_{\text{marg}}$, and lapse rate across groups and categories.
   - `likelihood_function`: Closure evaluating category likelihoods for arbitrary test observations across draws, returning a 3D array `[N_obs, K, N_draws]`.
   - `posterior_function`: Closure evaluating category posterior probabilities under specified decision rules, noise treatments, and lapse rates across draws, returning a 3D array `[N_obs, K, N_draws]`.
 - `summary`:
   - `parameters`: Point estimates of category parameters aggregated across draws.
   - `expected_moments`: Distribution summary (mean, SD, quantiles) of category expected moments.
   - `marginal_moments`: Distribution summary (mean, SD, quantiles) of category marginal moments.
   - `likelihood_function`: Closure evaluating category likelihoods directly from marginal moments (returning a 2D matrix `[N_obs, K]`).
   - `posterior_function`: Closure evaluating category posterior probabilities directly from marginal moments (returning a 2D matrix `[N_obs, K]`).

```{r add_posterior_latents_demo}
# Pre-extract posterior latents and closures into the model's @cache slot
fit <- add_posterior_latents(fit)

# Inspect cached quantities
names(fit@cache)
names(fit@cache$draws)
names(fit@cache$summary)
head(fit@cache$summary$expected_moments)
head(fit@cache$summary$marginal_moments)
```

### Why Precomputed Latents and Closures are Advantageous
1. **Performance**: Eliminates repeated data-wrangling loops during high-resolution contour or grid evaluations.
2. **Unified Evaluation**: Both draw-wise and summary evaluation closures share identical interfaces and handle noise marginalization and lapse rates automatically.
3. **Seamless Visualization**: Plotting functions such as `plot_categories()` and `plot_categorization_functions()` automatically leverage cached closures and moments.

For comprehensive examples of plotting category distributions and decision boundaries, see the companion vignette:
[Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html).

---

# 6. Model Comparison and Information Criteria {#model-comparison-and-criteria}

`MVBU_Stanfit` objects integrate with **loo** and **bayesplot** for Bayesian model comparison, cross-validation, and posterior predictive checking, following the argument conventions of **brms**.

## Adding Fit Criteria (`add_criterion`)

Model fit criteria (such as `"loo"`, `"waic"`, `"bayes_R2"`, `"loo_R2"`, `"kfold"`, `"loo_subsample"`, or `"marglik"`) are computed and stored in the model's `@criteria` slot using `add_criterion()`:

```{r add_criterion_demo}
# Compute and store LOO and WAIC criteria
fit <- add_criterion(fit, criterion = c("loo", "waic", "bayes_R2"), model_name = "NIW_2Cue_Model")

# Inspect stored criteria
names(fit@criteria)
fit@criteria$loo
fit@criteria$bayes_R2
```

## Model Comparison (`loo_compare`)

When evaluating alternative model specifications (e.g., comparing different prior specifications or model formulations), `loo_compare()` performs Leave-One-Out Cross-Validation model comparison via `loo::loo_compare()`:

```{r loo_compare_demo}
# Duplicate model for demonstration of loo_compare signature
fit2 <- fit
fit2@metadata$model_name <- "NIW_Alternative_Model"

# Compare models using LOO-CV
comp <- loo_compare(fit, fit2, criterion = "loo")
comp
```

> **Tip:** Leave-One-Out Cross-Validation (`loo`) is recommended over `waic` for robust model comparison, as `waic` can be sensitive to influential observations.

## Posterior Predictive Checking (`pp_check`)

Graphical posterior predictive checks compare observed exposure data $y$ with synthetic observations $y_{rep}$ drawn from the model's posterior predictive distribution:

```{r pp_check_demo}
if (requireNamespace("bayesplot", quietly = TRUE)) {
  # Overlay observed response density with posterior predictive replicates
  pp_check(fit, type = "dens_overlay", ndraws = 30)
}
```

---

# 7. Reconstructing Exposure Update Histories (`reconstruct_update_history`) {#reconstructing-exposure-update-histories}

When fitting an ideal adaptor model in Stan, researchers often wish to inspect how beliefs evolve incrementally across exposure observations.

The `reconstruct_update_history()` generic takes an `MVBU_Stanfit` object and reconstructs the sequential updating process into an `MVBU_ModelList` container:

- `uncertainty_treatment = "marginalize"` (default): Sequentially updates individual posterior MCMC draws (controlled by `ndraws`, default `20L`, with optional parallelization via `parallel = TRUE`). Returns an `MVBU_ModelList` holding lightweight `MVBU_StanfitPosterior` objects, faithfully capturing parameter estimation uncertainty across exposure checkpoints.
- `uncertainty_treatment = "discard"`: Fast mode that computes prior marginal moments and sequentially updates a single point-estimate ideal adaptor (e.g. `NIW_IdealAdaptor` or `NIX_IdealAdaptor`) across step checkpoints (e.g. `step_size = 10L`). Returns an `MVBU_ModelList` of `MVBU_CognitiveModel` objects.

```{r reconstruct_history_demo}
# 1. Fast point-estimate update sequence at checkpoints of 10 observations
model_seq_fast <- reconstruct_update_history(fit, uncertainty_treatment = "discard", step_size = 10L)
model_seq_fast

# 2. Full posterior update sequence with parameter uncertainty (10 draws)
model_seq_full <- reconstruct_update_history(fit, uncertainty_treatment = "marginalize", ndraws = 10L, step_size = 10L)
model_seq_full

# Inspect model labels along the exposure trajectory
head(model_seq_fast@model_labels)
```

The list of models that results from `reconstruct_update_history()` can be used as input to model visualizations, such a s `plot_categories` and `plot_categorization_functions`. See vignette "[Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html#visualizing-incremental-adaptation)" for examples of how to do this.

---

# 8. Summary and Related Documentation {#summary}

In summary, the `MVBU_Stanfit` workflow provides an integrated pipeline for Bayesian belief updating:

```
new_ideal_adaptor_stanfit_input() ──► fit_ideal_adaptor() ──► print() / summary()
                                                                      │
                                                                      ├──► add_posterior_latents() (Fast Plotting & Closures)
                                                                      ├──► add_criterion() & loo_compare() (Model Selection)
                                                                      └──► pp_check() (Predictive Checks)
```

### Links to Related Vignettes
- [S7 Class Architecture, Constructors, and Workflows](s7-class-structure-and-workflows.html): Detailed technical reference for S7 classes, generics, and model construction.
- [Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html): Comprehensive plotting guide for 1D, 2D, and 3D category distributions and decision boundaries.
- [Backward Compatibility: Working with Old Code](backward-compatibility-working-with-old-code.html): Transition guide from legacy list-column models to S7 architecture.
