---
title: "S7 Class Architecture, Constructors, and Workflows"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{S7 Class Architecture, Constructors, and Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..", quiet = TRUE)
}
```

<script>
document.addEventListener('DOMContentLoaded', function() {
  var tocLinks = document.querySelectorAll('#TOC a[href^="#"]');
  tocLinks.forEach(function(anchor) {
    anchor.addEventListener('click', function(e) {
      var href = this.getAttribute('href');
      var targetId = href.substring(1);
      var targetEl = document.getElementById(targetId) || 
                     document.getElementById('id_' + targetId) || 
                     document.querySelector('[name="' + targetId + '"]');
      if (targetEl) {
        e.preventDefault();
        targetEl.scrollIntoView({ behavior: 'smooth', block: 'start' });
        if (history.pushState) {
          history.pushState(null, null, href);
        }
      }
    });
  });
});
</script>

# 1. Introduction and The S7 Object Model {#introduction-and-s7-object-model}

The **MVBeliefUpdatr** package provides computational cognitive models of category representation, multidimensional perceptual decision making, and Bayesian belief updating. Version 1.0 of the package redesigns the underlying architecture around R's new **S7** object-oriented system (PEP / R Consortium object system).

In previous versions, category distributions and cognitive models were represented as nested tibbles with list-columns of vectors and matrices. While accessible for simple data-wrangling pipelines, weakly typed tabular models suffered from several limitations:

- Parameter matrices and dimensions were validated only at the point of computation, making errors hard to trace.
- Model metadata, cue labels, and decision rules were loosely coupled with distributional parameters.
- Method dispatch was split across ad-hoc S3 methods, legacy S4 classes, and family-specific wrapper functions.

The **S7 class architecture** replaces this with strongly typed, validated, and composable objects:

1. **Category Representations** (`MVBU_CategoryRepresentation`): Distributional beliefs about a single category in continuous multidimensional cue space (e.g., Gaussian [UVG/MVG/MUVG], Normal-Inverse-Wishart [NIW], Normal-Inverse-$\chi^2$ [NIX/MNIX], or exemplar sets).
2. **Category Representation Templates** (`MVBU_CategoryRepresentationTemplate`): Sets of category representations sharing a common cue space and coordinate system.
3. **Cognitive Decision Models** (`MVBU_CognitiveModel`): Complete observer or adaptor models composing a category template with category priors $P(c)$, decision rules (e.g., criterion, proportional, sampling), internal perceptual noise, and lapse behaviors.
4. **Stan Input & Fit Objects** (`MVBU_Staninput`, `MVBU_Stanfit`): Typed data bundles and Bayesian posterior fits for incremental belief-updating models.

For visualization workflows corresponding to these objects, see the companion vignette:
[Visualizing Categories, Categorization Functions, and Belief Updating](visualizing-models-and-categories.html).

For legacy code compatibility, see:
[Backward Compatibility: Working with Old Code](backward-compatibility-working-with-old-code.html).

```{r setup}
library(MVBeliefUpdatr)
library(tibble)
library(dplyr)
```

---

# 2. Quick Exploration with `example_*()` Functions {#quick-exploration-with-example-functions}

To immediately inspect the structure, labels, and parameters of any class in the package without manually assembling matrices, **MVBeliefUpdatr** provides a suite of lightweight `example_*()` constructors.

## Example Category Representations

Single-category representations can be instantiated for any supported family:

```{r example_representations}
# Multivariate Gaussian representation (2D: F1, F2)
rep_mvg <- example_category_representation("MVG")
rep_mvg

# Multivariate Univariate Gaussian representation (2D: independent variances)
rep_muvg <- example_category_representation("MUVG")
rep_muvg

# Normal-Inverse-Wishart belief distribution (2D)
rep_niw <- example_category_representation("NIW")
rep_niw

# Multivariate Normal-Inverse-Chi-Squared belief distribution (2D)
rep_mnix <- example_category_representation("MNIX")
rep_mnix

# Non-parametric Exemplar representation (2D)
rep_ex <- example_category_representation("EXEMPLAR")
rep_ex
```

Family-specific example helpers are also available, such as `example_uvg_category_representation()`, `example_mvg_category_representation()`, `example_muvg_category_representation()`, `example_niw_category_representation()`, `example_mnix_category_representation()`, and `example_exemplar_category_representation()`.

## Example Category Representation Templates

Templates bind multiple category representations together into a shared space:

```{r example_templates}
# 2-category vowel template (/i/, /u/)
tpl_mvg <- example_category_representation_template("MVG")
tpl_mvg

# Inspect the categories and cues in the template
get_category_labels(tpl_mvg)
get_cue_labels(tpl_mvg)
```

## Example Cognitive Decision Models

Full decision models can be generated directly across all supported model families:

```{r example_models}
# Ideal Observer with MVG category representations
mod_observer <- example_mvg_ideal_observer()
mod_observer

# Ideal Observer with MUVG category representations
mod_muvg <- example_muvg_ideal_observer()
mod_muvg

# Ideal Adaptor with NIW category representations
mod_adaptor <- example_niw_ideal_adaptor()
mod_adaptor

# Ideal Adaptor with MNIX category representations
mod_mnix <- example_mnix_ideal_adaptor()
mod_mnix

# Exemplar decision model
mod_exemplar <- example_exemplar_model()
mod_exemplar
```

## Example Fitted Stan Models

Fitted incremental belief-updating models (`MVBU_Stanfit`) can be accessed through `get_example_stanfit()`:

```{r example_stanfit}
# Load cached 1-cue NIW ideal adaptor stanfit
fit_1cue <- get_example_stanfit(1, stanmodel = "NIW_ideal_adaptor", file_refit = "never")
fit_1cue
```

---

# 3. Object Hierarchy and Constructors {#object-hierarchy-and-constructors}

The class system is organized into three hierarchical levels:

```
MVBU_Object
  ├── MVBU_CategoryRepresentation
  │     ├── UVG_CategoryRepresentation
  │     ├── MVG_CategoryRepresentation
  │     ├── MUVG_CategoryRepresentation
  │     ├── NIW_CategoryRepresentation
  │     ├── NIX_CategoryRepresentation
  │     ├── MNIX_CategoryRepresentation
  │     └── Exemplar_CategoryRepresentation
  ├── MVBU_CategoryRepresentationTemplate
  ├── MVBU_CognitiveModel
  │     ├── MVG_IdealObserver
  │     ├── MUVG_IdealObserver
  │     ├── NIW_IdealAdaptor
  │     ├── MNIX_IdealAdaptor
  │     └── Exemplar_Model
  ├── MVBU_Staninput
  │     └── IdealAdaptorStaninput
  │           ├── NIX_IdealAdaptorStaninput
  │           └── NIW_IdealAdaptorStaninput
  └── MVBU_Stanfit
        └── IdealAdaptorStanfit
              ├── NIX_IdealAdaptorStanfit
              └── NIW_IdealAdaptorStanfit
```

## Level 1: Category Representations {#level-1-category-representations}

### Low-Level Bare Constructors

When category parameters are already known, representations can be constructed by passing numeric vectors and matrices directly:

```{r bare_representation_constructors}
# Multivariate Gaussian: mean vector mu and covariance matrix Sigma
mu_b <- c(F1 = 350, F2 = 1100)
Sigma_b <- matrix(c(2500, 500, 500, 3600), nrow = 2, ncol = 2,
                  dimnames = list(names(mu_b), names(mu_b)))

rep_b <- new_mvg_category_representation(
  category_labels = "/b/",
  cue_labels = c("F1", "F2"),
  mu = mu_b,
  Sigma = Sigma_b
)
rep_b

# MUVG: Multivariate Univariate Gaussian (diagonal variances sigma2)
rep_muvg_custom <- new_muvg_category_representation(
  category_labels = "/b/",
  cue_labels = c("F1", "F2"),
  component_mu = mu_b,
  component_sigma2 = c(F1 = 2500, F2 = 3600)
)
rep_muvg_custom
```

For conjugate prior representations like NIW or MNIX, specify prior parameter moments:

```{r bare_niw_constructor}
rep_niw_custom <- new_niw_category_representation(
  category_labels = "/b/",
  cue_labels = c("F1", "F2"),
  m = c(F1 = 350, F2 = 1100),
  S = matrix(c(25000, 5000, 5000, 36000), nrow = 2, ncol = 2),
  kappa = 10,
  nu = 12
)
rep_niw_custom

# MNIX (Multivariate Normal-Inverse-Chi-Squared for independent cues)
rep_mnix_custom <- new_mnix_category_representation(
  category_labels = "/b/",
  cue_labels = c("F1", "F2"),
  component_m = c(F1 = 350, F2 = 1100),
  component_sigma2 = c(F1 = 2500, F2 = 3600),
  component_kappa = c(F1 = 10, F2 = 10),
  component_nu = c(F1 = 12, F2 = 12)
)
rep_mnix_custom
```

### Empirical Constructors from Data

When starting from empirical observations, use `new_category_representation_from_data()` (or family-specific constructors like `new_mvg_category_representation_from_data()`):

```{r data_representation_constructors}
# Sample acoustic data for a category
set.seed(42)
obs_data <- data.frame(
  F1 = rnorm(50, mean = 400, sd = 40),
  F2 = rnorm(50, mean = 1800, sd = 120),
  vowel = "ih"
)

# Estimate MVG representation from observations
rep_from_data <- new_mvg_category_representation_from_data(
  data = obs_data,
  category = "vowel",
  cues = c("F1", "F2")
)
rep_from_data
```

## Level 2: Category Representation Templates {#level-2-category-representation-templates}

Templates aggregate multiple category representations into a unified inventory.

### Manual Template Construction

```{r manual_template}
mu_d <- c(F1 = 450, F2 = 1700)
Sigma_d <- matrix(c(3000, 200, 200, 4000), nrow = 2, ncol = 2,
                  dimnames = list(names(mu_d), names(mu_d)))
rep_d <- new_mvg_category_representation(
  category_labels = "/d/",
  cue_labels = c("F1", "F2"),
  mu = mu_d,
  Sigma = Sigma_d
)

# Combine representations into a template
tpl_custom <- new_category_representation_template(
  representations = list(rep_b, rep_d)
)
tpl_custom
```

### Empirical Template from Data

```{r empirical_template}
# Multi-category training data
train_data <- data.frame(
  F1 = c(rnorm(40, 300, 30), rnorm(40, 600, 40)),
  F2 = c(rnorm(40, 2200, 80), rnorm(40, 1100, 70)),
  category = rep(c("/i/", "/u/"), each = 40)
)

tpl_empirical <- new_mvg_category_representation_template_from_data(
  data = train_data,
  category = "category",
  cues = c("F1", "F2")
)
tpl_empirical
```

## Level 3: Cognitive Decision Models {#level-3-cognitive-decision-models}

A cognitive model combines a category template with:

- `category_prior`: Prior probabilities $P(c)$ across categories (must sum to 1).
- `decision_rule`: How posterior probabilities map to categorization responses (`"criterion"`, `"proportional"`, or `"sampling"`).
- `lapse_rate` and `lapse_bias`: Probability and distribution of lapse responses (unrelated to cue values).
- `Sigma_noise` and `noise_treatment`: Internal perceptual noise covariance and handling (`"no_noise"`, `"sample"`, or `"marginalize"`).

```{r cognitive_model_constructors}
# Instantiate an Ideal Observer model from the template
observer_model <- new_mvg_ideal_observer(
  category_template = tpl_empirical,
  category_prior = c("/i/" = 0.5, "/u/" = 0.5),
  decision_rule = "proportional",
  lapse_rate = 0.02
)
observer_model
```

Similarly, empirical constructors can build the complete cognitive model directly from data in a single step:

```{r model_from_data}
model_from_data <- new_mvg_ideal_observer_from_data(
  data = train_data,
  category = "category",
  cues = c("F1", "F2"),
  decision_rule = "criterion"
)
model_from_data
```

---

# 4. Standard Generics: Inspection, Evaluation, and Sampling {#standard-generics}

The S7 architecture provides consistent generic functions across all model families.

## Label Accessors

```{r label_accessors}
# Extract cue names
get_cue_labels(observer_model)

# Extract category names
get_category_labels(observer_model)

# Extract combined label structure
get_labels(observer_model)
```

## Type and Family Introspection

```{r introspection}
# Model family
get_model_type(observer_model)

# Representation family
get_representation_type(rep_b)
```

## Parameter Extraction

Extract category parameters or expected moments across representations, templates, models, and fitted Bayesian models:

```{r parameter_extraction}
# Extract parameters across template categories (mu, Sigma for MVG)
get_parameters(observer_model)

# For Bayesian belief distributions (like the NIW ideal adaptor),
# extract expected moments under the prior distribution:
get_expected_mu(mod_adaptor)
get_expected_sigma(mod_adaptor)
```

## Likelihood and Posterior Computation

Evaluate category likelihoods $p(x \mid c)$ and posterior probabilities $P(c \mid x)$ for new observations:

```{r likelihood_and_posterior}
# New test observations in cue space
test_points <- matrix(
  c(320, 2100,
    580, 1150),
  nrow = 2, byrow = TRUE,
  dimnames = list(NULL, c("F1", "F2"))
)

# Compute category likelihoods
lik_mat <- likelihood(observer_model, test_points)
lik_mat

# Compute posterior category probabilities P(c | x)
post_mat <- posterior(observer_model, test_points)
post_mat
```

## Categorization Under Decision Rules

First, let's inspect the configured decision rule of our cognitive model:

```{r check_decision_rule}
# Inspect the configured decision rule of the model
observer_model@decision_rule
```

Now call `categorize()` to apply the model's decision rule to the test observations:

```{r categorize}
# Categorize test points using the model's configured decision rule
cat_results <- categorize(observer_model, test_points)
cat_results

# Override decision rule to deterministic criterion ("winner-take-all")
categorize(observer_model, test_points, decision_rule = "criterion")
```

### Understanding the `probability` Column in `categorize()`

The `probability` column in the data frame returned by `categorize()` reports the **posterior category probability** $P(c \mid x)$ of the selected category, calculated directly from the model's posterior belief distribution:

$$P(c \mid x) = \frac{p(x \mid c) \, P(c)}{\sum_{c'} p(x \mid c') \, P(c')}$$

It does **not** represent the *response choice probability* under the decision rule. For example, under a deterministic decision rule like `'criterion'`, the model always selects the MAP category with response probability $1$, but the `probability` column displays $P(c_{\text{MAP}} \mid x) = 0.94$, indicating the observer's degree of posterior certainty in that classification decision.

## Sampling Observations

Generate synthetic perceptual tokens drawn from category representations:

```{r sample_observations}
# Sample 5 observations from the /i/ category representation
sample_observations(tpl_empirical@representations[["/i/"]], n = 5)

# Sample 6 observations across categories weighted by category priors
sample_observations(observer_model, n = 6)
```

---

# 5. Bayesian Fitting Architecture: `Staninput` and `Stanfit` {#bayesian-fitting-architecture}

Fitting ideal adaptor models with incremental belief updating (IBBU) follows a clean two-tier pipeline.

```
Observations (data.frame)
        │
        ▼
prepare_staninput() ───► IdealAdaptorStanfitInput (S7)
                                 │
                                 ▼
                         fit() / stan()
                                 │
                                 ▼
                         MVBU_Stanfit (S7)
```

## Preparing Typed Stan Input (`prepare_staninput`)

`prepare_staninput()` transforms empirical exposure data, priors, and cue specifications into a typed S7 `IdealAdaptorStanfitInput` object:

```{r prepare_staninput, eval = FALSE}
# (Workflow demonstration; requires model specification)
stan_input <- prepare_staninput(
  data = train_data,
  cues = c("F1", "F2"),
  category_col = "category",
  stanmodel = "NIW_ideal_adaptor"
)

# Inspect typed staninput slots
class(stan_input)
get_cue_labels(stan_input)
get_transform_information(stan_input)
```

## Inspecting Fitted Stan Models (`MVBU_Stanfit`)

Fitted models encapsulate the Stan MCMC output, typed Stan input, cue transform metadata, and prior parameters:

```{r stanfit_inspection}
# Inspect example fitted model
fit_1cue
get_model_type(fit_1cue)

# Extract underlying rstan stanfit object
raw_stanfit <- get_stanfit(fit_1cue)
class(raw_stanfit)

# Extract MCMC draws in tidy format
draws_df <- get_draws(fit_1cue)
head(draws_df)
```

## Posterior Draws and Diagnostics

`MVBU_Stanfit` supports standard diagnostics and posterior formats from the **posterior**, **bayesplot**, and **loo** ecosystems:

```{r stanfit_diagnostics}
# Rhat convergence diagnostic
head(rhat(fit_1cue))

# Effective sample size ratio
head(neff_ratio(fit_1cue))

# Model summary
summary(fit_1cue)
```

---

# 6. Aggregating Models and Representations {#aggregating-models}

In many experimental and modeling scenarios (e.g., combining individual-participant models or creating group-level representations), multiple cognitive models or category templates need to be aggregated into a single combined model. The `aggregate_models()` function and `aggregate()` generic handle this across S7 classes.

## Aggregating Multivariate Gaussian (MVG) Models

Consider three individual observer models representing vowel categories with distinctly different estimated means and covariances (e.g., positive correlation, negative correlation, and uncorrelated cues):

```{r mvg_aggregation}
library(ggplot2)
library(patchwork)

# Individual Observer 1: Positive cue covariance, shifted top-left
mvg1 <- new_mvg_ideal_observer(
  category_template = new_category_representation_template(
    representations = list(
      "/i/" = new_mvg_category_representation("/i/", c("F1", "F2"), mu = c(250, 2400), Sigma = matrix(c(800, 600, 600, 3000), 2, 2, dimnames = list(c("F1", "F2"), c("F1", "F2")))),
      "/u/" = new_mvg_category_representation("/u/", c("F1", "F2"), mu = c(420, 1200), Sigma = matrix(c(900, 500, 500, 2500), 2, 2, dimnames = list(c("F1", "F2"), c("F1", "F2"))))
    )
  ),
  lapse_rate = 0.01
)

# Individual Observer 2: Negative cue covariance and larger variance, shifted right
mvg2 <- new_mvg_ideal_observer(
  category_template = new_category_representation_template(
    representations = list(
      "/i/" = new_mvg_category_representation("/i/", c("F1", "F2"), mu = c(350, 2200), Sigma = matrix(c(2000, -1200, -1200, 6000), 2, 2, dimnames = list(c("F1", "F2"), c("F1", "F2")))),
      "/u/" = new_mvg_category_representation("/u/", c("F1", "F2"), mu = c(580, 1000), Sigma = matrix(c(1800, -800, -800, 4500), 2, 2, dimnames = list(c("F1", "F2"), c("F1", "F2"))))
    )
  ),
  lapse_rate = 0.05
)

# Individual Observer 3: Uncorrelated cues, shifted down
mvg3 <- new_mvg_ideal_observer(
  category_template = new_category_representation_template(
    representations = list(
      "/i/" = new_mvg_category_representation("/i/", c("F1", "F2"), mu = c(300, 2000), Sigma = matrix(c(1200, 0, 0, 4000), 2, 2, dimnames = list(c("F1", "F2"), c("F1", "F2")))),
      "/u/" = new_mvg_category_representation("/u/", c("F1", "F2"), mu = c(500, 800), Sigma = matrix(c(1500, 0, 0, 3500), 2, 2, dimnames = list(c("F1", "F2"), c("F1", "F2"))))
    )
  ),
  lapse_rate = 0.03
)

# Inspect individual models before aggregation
cat("--- Individual Model 1 ---\n")
mvg1
cat("\n--- Individual Model 2 ---\n")
mvg2
cat("\n--- Individual Model 3 ---\n")
mvg3

# Aggregate all three models (unweighted mean)
mvg_aggregated <- aggregate_models(list(mvg1, mvg2, mvg3))

cat("\n--- Aggregated MVG Model ---\n")
mvg_aggregated
```

To see how category representation parameters change during aggregation, we can explicitly compare the category means ($\mu$) and covariance matrices ($\Sigma$) before and after aggregation:

```{r mvg_aggregation_comparison}
# Compare category parameters before and after aggregation
for (cat_name in c("/i/", "/u/")) {
  cat(sprintf("\n=== Category: %s ===\n", cat_name))
  cat("Model 1 mu:", mvg1@category_template@representations[[cat_name]]@mu, "\n")
  cat("Model 2 mu:", mvg2@category_template@representations[[cat_name]]@mu, "\n")
  cat("Model 3 mu:", mvg3@category_template@representations[[cat_name]]@mu, "\n")
  cat("Aggregated mu:", mvg_aggregated@category_template@representations[[cat_name]]@mu, "\n\n")
  
  cat("Model 1 Sigma:\n")
  print(mvg1@category_template@representations[[cat_name]]@Sigma)
  cat("Model 2 Sigma:\n")
  print(mvg2@category_template@representations[[cat_name]]@Sigma)
  cat("Model 3 Sigma:\n")
  print(mvg3@category_template@representations[[cat_name]]@Sigma)
  cat("Aggregated Sigma:\n")
  print(mvg_aggregated@category_template@representations[[cat_name]]@Sigma)
}
```

We can also visualize the category likelihood ellipses of the individual models alongside the aggregated model:

```{r mvg_aggregation_plots, fig.width=9, fig.height=7, echo=TRUE}
p1 <- plot(mvg1) + ggtitle("Model 1 (Pos Corr)")
p2 <- plot(mvg2) + ggtitle("Model 2 (Neg Corr)")
p3 <- plot(mvg3) + ggtitle("Model 3 (Uncorrelated)")
p_agg <- plot(mvg_aggregated) + ggtitle("Aggregated Model (Average)")

(p1 + p2) / (p3 + p_agg)
```



## Aggregating Exemplar Models

Exemplar models are aggregated by concatenating their underlying exemplar sets, normalizing exemplar weights, and averaging lapse rates and tuning parameters:

```{r exemplar_aggregation}
# Create three exemplar models with distinct stored exemplars
ex_data1 <- matrix(c(300, 2200, 310, 2180), ncol = 2, dimnames = list(NULL, c("F1", "F2")))
ex_data2 <- matrix(c(320, 2150, 325, 2130), ncol = 2, dimnames = list(NULL, c("F1", "F2")))
ex_data3 <- matrix(c(280, 2250, 290, 2230), ncol = 2, dimnames = list(NULL, c("F1", "F2")))

ex_m1 <- new_exemplar_model(
  category_template = new_category_representation_template(
    representations = list(
      "/i/" = new_exemplar_category_representation("/i/", c("F1", "F2"), exemplars = ex_data1, c = 1.0)
    )
  ),
  lapse_rate = 0.02
)

ex_m2 <- new_exemplar_model(
  category_template = new_category_representation_template(
    representations = list(
      "/i/" = new_exemplar_category_representation("/i/", c("F1", "F2"), exemplars = ex_data2, c = 1.5)
    )
  ),
  lapse_rate = 0.04
)

ex_m3 <- new_exemplar_model(
  category_template = new_category_representation_template(
    representations = list(
      "/i/" = new_exemplar_category_representation("/i/", c("F1", "F2"), exemplars = ex_data3, c = 2.0)
    )
  ),
  lapse_rate = 0.03
)

# Inspect individual exemplar models before aggregation
cat("--- Individual Exemplar Model 1 ---\n")
ex_m1
cat("\n--- Individual Exemplar Model 2 ---\n")
ex_m2
cat("\n--- Individual Exemplar Model 3 ---\n")
ex_m3

# Aggregate all three exemplar models
ex_aggregated <- aggregate_models(list(ex_m1, ex_m2, ex_m3))

cat("\n--- Aggregated Exemplar Model ---\n")
ex_aggregated

# Total combined exemplars stored in the aggregated model:
nrow(ex_aggregated@category_template@representations[["/i/"]]@exemplars)
```
