---
title: "Visualizing Categories, Categorization Functions, and Belief Updating"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Visualizing Categories, Categorization Functions, and Belief Updating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all("..", quiet = TRUE)
}
```

<script>
document.addEventListener('DOMContentLoaded', function() {
  var tocLinks = document.querySelectorAll('#TOC a[href^="#"]');
  tocLinks.forEach(function(anchor) {
    anchor.addEventListener('click', function(e) {
      var href = this.getAttribute('href');
      var targetId = href.substring(1);
      var targetEl = document.getElementById(targetId) || 
                     document.getElementById('id_' + targetId) || 
                     document.querySelector('[name="' + targetId + '"]');
      if (targetEl) {
        e.preventDefault();
        targetEl.scrollIntoView({ behavior: 'smooth', block: 'start' });
        if (history.pushState) {
          history.pushState(null, null, href);
        }
      }
    });
  });
});
</script>

# Introduction

Visualizing distributional representations and response probabilities is essential for understanding multidimensional category structures, decision boundaries, and belief updating. The **MVBeliefUpdatr** package provides a unified, highly extensible visualization system built on top of **ggplot2** and **plotly**. Any of the returned `ggplot` or `plotly` objects can be customized with any native `ggplot2` or `plotly` layers, color scales, facets, and themes.

Every visualization function in the package dispatches on the new **S7** class architecture, accepting:

- Single category representations (`MVBU_CategoryRepresentation`)
- Multi-category templates (`MVBU_CategoryRepresentationTemplate`)
- Complete cognitive decision models (`MVBU_CognitiveModel`)
- Fitted Bayesian models (`MVBU_Stanfit`)

```{r setup}
library(MVBeliefUpdatr)
library(ggplot2)
library(patchwork)
library(plotly)
```

---

# 1. Visualizing Category Representations (`plot_categories`) {#visualizing-category-representations}

The `plot_categories()` generic visualizes the distributional layout of category beliefs in cue space. Users can select which cues to plot through the `cues` argument:

 - `cues`: Character vector of cues to be plotted. By default, as many cues as contained in the category representation(s), up to a maximum of three cues. If there are more than three cues, the first three cues will be plotted. Selecting fewer than the full set of cues causes the function to [marginalize over the unplotted cues](#dimensionality-reduction-and-marginal-projections).
 - `categories`: Optional character vector subset of category labels to plot (e.g., `categories = c("/b/", "/d/")`). Defaults to all categories.
 
## 1D Representations and Templates {#1d-representations-and-templates}

In one cue is selected for plotting (`length(cues) == 1L`), category representations are visualized as univariate probability density functions $p(x \mid c)$. Core arguments (listed in order of function signature):

 - `aes`: Aesthetic layer type (`"contour"` for density line curves, `"fill-discrete"` or alias `"fill"` for shaded density ribbons; `"fill-gradient"` yields equivalent continuous density shading).
 - `limits`: A 2-element numeric vector `c(min, max)` or named list `list(cue = c(min, max))` specifying axis bounds. Defaults to the span of the plotted distributions across all categories.
 - `resolution`: Number of evaluation points along the continuous cue grid (default `200`).

Other arguments are ignored.

```{r 1d_categories_mvg}
# Construct a 1D MVG category representation for /b/ along Voice-Onset Time (ms)
rep_1d_mvg <- example_mvg_category_representation(n_cues = 1, category = "/b/")
p_1d_rep <- plot_categories(rep_1d_mvg)
p_1d_rep
```

When applied to a multi-category template or cognitive model, all categories are plotted simultaneously with unified color mapping and a shared legend:

```{r 1d_categories_template, fig.width = 8, fig.height = 4}
# 1D MVG multi-category template (/b/, /d/, /g/)
tpl_1d_mvg <- example_mvg_category_representation_template(n_cues = 1, categories = c("/b/", "/d/", "/g/"))

# Left: Line contour curves (default aes = 'contour')
p_tpl_line <- plot_categories(tpl_1d_mvg, limits = c(-50, 120)) +
  labs(title = "Line Density (default aes = 'contour')")

# Right: Shaded density ribbons (aes = 'fill')
p_tpl_fill <- plot_categories(tpl_1d_mvg, aes = "fill", limits = c(-50, 120)) +
  labs(title = "Shaded Ribbon (aes = 'fill')")

p_tpl_line + p_tpl_fill + plot_layout(guides = "collect")
```

## 1D Exemplar-Based Categories {#exemplar-based-categories-and-sampling-control}
For exemplar representations (`Exemplar_CategoryRepresentation`), `plot_categories()` evaluates a non-parametric kernel density estimate (KDE) over the stored discrete exemplars.

 - `n_exemplars`: Number of stored exemplars to sample and display as rug tick marks (default `0L`, plotting clean density curves without rugs). When `n_exemplars > 0`, the rug tick marks are rendered with an opacity of $1 / K$ (where $K$ is the number of plotted categories).

```{r 1d_categories_exemplar}
tpl_1d_ex <- example_exemplar_category_representation_template(n_cues = 1)

# Sampling 100 stored exemplars to render rug tick marks
plot_categories(tpl_1d_ex, n_exemplars = 100)
```

---

## Customizing the Returned `ggplot` or `plotly` Object {#customizing-the-returned-ggplot-object}
As mentioned above, all plotting functions return standard `ggplot` or `plotly` objects. So you can customize them with any native `ggplot2` or `plotly` layers, color scales, facets, and themes. For instance, for the 1d category representation plot, we could choose a theme, add labs, and further annotation, and change the color assignment:

```{r 1d_ggplot_customization}
# Customizing scales, adding reference thresholds and annotations
p_tpl_line +
  scale_color_manual(values = c("/b/" = "#E41A1C", "/d/" = "#377EB8", "/g/" = "#4DAF4A")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Voice-Onset Time Distributions across Stop Consonants",
    subtitle = "Parametric Gaussian representations with custom aesthetics",
    x = "Voice-Onset Time (ms)",
    y = "Probability Density"
  ) +
  theme_minimal()
```

---

## 2D Representations and Templates {#2d-category-plots-aesthetics-and-layer-modes}

When two cues are selected for plotting (`length(cues) == 2L`), category representations can be visualized using covariance ellipses, continuous density gradients, or discrete density bands. Core arguments (in order of function signature):

 - `aes`: Aesthetic layer type:
   - `"contour"`: Draws boundary contour lines (covariance ellipses for parametric models; 2D KDE contours for exemplar models).
   - `"fill-discrete"` (alias `"fill"`): Shaded discrete confidence regions colored by category.
   - `"fill-gradient"`: Continuous density heatmap with alpha proportional to $p(\mathbf{x} \mid c)$.
   - Combinations like `c("fill-gradient", "contour")` or `c("fill-discrete", "contour")`. When both fill and contour aesthetics are present, their fill and color keys are automatically unified into a single legend.
 - `levels`: Central probability mass regions to enclose (defaulting to `levels = 2 * pnorm(1:3) - 1` $\approx 0.683, 0.954, 0.997$). Users can specify custom regions, such as `levels = c(0.50, 0.95)` for 50% and 95% intervals. 
 - `limits`: Coordinate bounds along the axes. Specified as a named list (e.g., `limits = list(VOT = c(-30, 90), f0_semitones = c(-4, 6))`) or a 2-element numeric vector `c(min, max)` applied to both dimensions. When `"fill-gradient"` is used, the density raster tiles seamlessly cover the entire plotted space up to the panel boundaries without artificial padding.
 - `resolution`: Grid density along each cue axis (default `100`). 

```{r 2d_categories_mvg_modes, fig.width = 9.5, fig.height = 4}
tpl_2d_mvg <- example_mvg_category_representation_template(n_cues = 2)

# Mode 1: Custom 50% and 95% central probability contours for /b/ and /p/
p_2d_contour <- plot_categories(
  tpl_2d_mvg, 
  categories = c("/b/", "/p/"),
  aes = "contour", 
  levels = c(0.50, 0.95),
  limits = list(VOT = c(-30, 90), f0_semitones = c(-4, 6))
) + labs(title = "50% & 95% Central Contours")

# Mode 2: Continuous density gradient + contour borders for /b/ and /p/
p_2d_combined <- plot_categories(
  tpl_2d_mvg, 
  categories = c("/b/", "/p/"),
  aes = c("fill-gradient", "contour"),
  limits = list(VOT = c(-30, 90), f0_semitones = c(-4, 6))
) + labs(title = "Gradient Fill + Contours")

p_2d_contour + p_2d_combined + plot_layout(guides = "collect")
```

## 2D Exemplar Models: Raw Exemplars vs. KDE Densities {#2d-exemplar-models-raw-exemplars-vs-kde-densities}
For 2D exemplar templates, `plot_categories()` evaluates vectorized bivariate kernel densities and optionally overlays sampled exemplar points:

```{r 2d_categories_exemplar, fig.width = 7, fig.height = 5}
tpl_2d_ex <- example_exemplar_category_representation_template(n_cues = 2)

# Plot 2D KDE density contours and sample 120 exemplars as scatter points
plot_categories(
  tpl_2d_ex, 
  categories = c("/b/", "/p/"),
  aes = "contour", 
  n_exemplars = 120,
  limits = list(VOT = c(-30, 90), f0_semitones = c(-4, 6))
)
```

## Interactive Categories over 2D cues: 3D Density Surfaces {#2d-interactive-category-plots-3d-density-surfaces}

When `interactive = TRUE` (or `aes = "interactive"`) is set for a 2-cue plot, `plot_categories()` renders an interactive 3D WebGL scene via **plotly** where the third vertical dimension represents probability density $p(x_1, x_2 \mid c)$:

 - `aes`: Aesthetic surface mode:
   - `"fill-discrete"` (alias `"fill"`, default): 3D density surface with solid category coloring and constant opaqueness across the surface.
   - `"fill-gradient"`: 3D density surface with opacity gradient proportional to the height of the probability density.
   - `"contour"`: Only 3D contour lines across density heights (hiding the surface fill).

```{r 2d_interactive_demo, fig.width = 8, fig.height = 6}
# Interactive 3D density surface over two acoustic cues (default solid fill-discrete)
plot_categories(
  tpl_2d_mvg,
  categories = c("/b/", "/p/"),
  interactive = TRUE
)
```

Just like the `ggplot` objects returned by non-interactive 1D and 2D plots can be further customized using standard `ggplot2` functions, so too can the **plotly** objects returned by interactive 2D and 3D plots be customized using **plotly** functions (e.g., using `layout()` to adjust the camera angle or aspect ratio). 

---

## 3D Visualizations of Category Representations {#3d-visualizations-of-category-representations}

When working with three cues (such as $F_1 \times F_2 \times F_3$ in vowel perception or $\text{VOT} \times f_0 \times \text{duration}$ in stop voicing), categories inhabit a volume in $\mathbb{R}^3$. **MVBeliefUpdatr** supports two complementary approaches: multi-panel 2D slices and fully interactive 3D WebGL scenes.

### Sliced Multi-Panel Visualizations {#sliced-multi-panel-visualizations}

The sliced view conditions on discrete values of the third cue, rendering a multi-panel `ggplot` figure where each panel displays the 2D subspace at that slice level:

 - `cues`: Character vector of length 3 `c("cue1", "cue2", "cue3")`, where the third cue is sliced across facets.
 - `slices`: Numeric vector of values along the third cue to slice at (e.g. `slices = c(2200, 2600, 3000)`). If omitted, slices default to 5 standard deviation intervals at $-2, -1, 0, 1,$ and $2\sigma$ from the pooled mean of the third cue.
 - `levels`: Central probability mass intervals (defaults to 1 to 4 standard deviation equivalents: $\approx 0.683, 0.954, 0.997, 0.9999$). Each contour line displays a text label indicating the percentage of probability mass it encloses, with line width varying linearly from thickest at the central mass to thinnest at the outermost boundary.

```{r 3d_sliced_categories, fig.width = 9.5, fig.height = 4}
# Construct a 3-cue MVG Ideal Observer
m3 <- new_mvg_ideal_observer(
  category_template = new_category_representation_template(list(
    b = new_mvg_category_representation("b", c("F1", "F2", "F3"), c(300, 1000, 2200), diag(c(40, 120, 180)^2)),
    d = new_mvg_category_representation("d", c("F1", "F2", "F3"), c(450, 1600, 2600), diag(c(50, 140, 200)^2)),
    g = new_mvg_category_representation("g", c("F1", "F2", "F3"), c(600, 2200, 3000), diag(c(55, 160, 220)^2))
  )),
  category_prior = c(b = 0.33, d = 0.33, g = 0.34)
)

# Sliced 3D category plot using default 5 slices (-2 to +2 sigma) and default 1:3 sigma contours
plot_categories(m3)
```

---

### 3D Interactive Category Plots: (Plotly) {#interactive-3d-webgl-plots-plotly}

For interactive 3D exploration, `plot_categories()` provides native WebGL rendering via **plotly** simply by setting `interactive = TRUE` (or specifying `aes = "interactive"`):

 - `levels`: Central probability mass intervals. Defaults to the 2-sigma central ellipsoid ($\approx 95.45\%$).
 - `aes`: Aesthetic layer type, the default of which depends on the type of category representation:
   - `"fill-discrete"` (alias `"fill"`, default for parametric models): 3D ellipsoids with constant semi-opacity per category.
   - `"fill-gradient"`: Nested ellipsoids with opacity gradient across probability mass levels.
   - `"contour"`: 3D wireframe contour lines.
   - `"scatter"`: For exemplar models only (default), interactive 3D scatter point cloud (controlled by `n_exemplars`). 

While Exemplar models can also use `aes = "fill-discrete"`, `"fill-gradient"`, or `"contour"`, note that **3D isosurface generation for exemplar models is computationally intensive.**

```{r 3d_interactive_demo, fig.width = 8, fig.height = 6}
# Native interactive 3D WebGL category plot with default 2-sigma ellipsoid
plot_categories(
  m3, 
  interactive = TRUE
)
```

Interactive 3D plots are fully responsive: users can click and drag to rotate the 3D volume, scroll to zoom, hover to inspect category labels, and toggle category visibilities in the legend.

### 3D Exemplar Models: Sliced Contours vs. Interactive Scatter Clouds {#3d-exemplar-models-sliced-contours-vs-interactive-scatter-clouds}
For 3D exemplar templates, `plot_categories()` provides two distinct visualization strategies:

1. **Sliced 2D Density Contours (`interactive = FALSE`)**: Evaluates bivariate Kernel Density Estimation (KDE) over 2D slices along the third cue dimension, displaying how the local exemplar distribution and category densities shift across cue values:

```{r 3d_exemplar_sliced, fig.width = 9.5, fig.height = 4}
tpl_3d_ex <- example_exemplar_category_representation_template(n_cues = 3)

# Sliced 2D density contours for /b/ and /p/ across vowel_duration slices (50, 100, 150 ms)
plot_categories(
  tpl_3d_ex, 
  categories = c("/b/", "/p/"),
  slices = c(50, 100, 150),
  resolution = 35
)
```

2. **Interactive 3D Scatter Clouds (`interactive = TRUE`)**: Renders stored exemplars directly as a 3D point cloud:

```{r 3d_exemplar_interactive, fig.width = 8, fig.height = 6}
# Interactive 3D scatter plot sampling 150 exemplars per category (overriding default 100)
plot_categories(
  tpl_3d_ex, 
  categories = c("/b/", "/p/"),
  interactive = TRUE, 
  n_exemplars = 150
)
```

*Note on continuous 3D density surfaces for exemplar models:* While parametric representations (e.g., MVG) have closed-form ellipsoid equations that render efficiently in 3D WebGL, continuous 3D density isosurfaces for non-parametric exemplar models require estimating a 3D kernel density across a high-resolution 3D voxel grid ($O(N \cdot G^3)$ evaluations, where $N$ can exceed 50,000 empirical tokens) followed by 3D isosurface mesh extraction (e.g., marching cubes). In addition to being computationally demanding, nested 3D density shells suffer from severe visual occlusion in complex multi-category spaces. Consequently, sliced 2D contour panels and interactive 3D scatter clouds provide the clearest, fastest, and most faithful visual diagnostics for 3D exemplar categories.

---

# 2. Categorization Functions (`plot_categorization_function`) {#categorization-functions}

While `plot_categories()` displays generative likelihood densities $p(\mathbf{x} \mid c)$, `plot_categorization_function()` visualizes predicted response probabilities $P_{\text{resp}}(c \mid \mathbf{x})$, incorporating perceptual decision rules, lapse rates, and category biases:
$$P_{\text{resp}}(c \mid \mathbf{x}) = (1 - \lambda) P(c \mid \mathbf{x}) + \lambda b_c$$
where:

 - $\lambda \in [0, 1]$ is the **lapse rate** (representing probability of random guessing, perceptual error, or task-unrelated lapses of attention).
 - $b_c \in [0, 1]$ is the **lapse bias** for category $c$, with $\sum_{k} b_k = 1$.
 - $P(c \mid \mathbf{x})$ is the posterior response probability derived from the generative model:
   $$P(c \mid \mathbf{x}) = \frac{p(\mathbf{x} \mid c) P(c)}{\sum_{k} p(\mathbf{x} \mid k) P(k)}$$
   according to the chosen `decision_rule` (e.g., `"proportional"` for probability matching or `"criterion"` for deterministic winner-take-all classification).

*For an introduction to lapse rates, category biases, and perceptual noise in speech categorization models, see [Xie, Jaeger, & Kurumada (2023)](https://doi.org/10.1016/j.cortex.2023.05.011).* Core Arguments for `plot_categorization_function` include:

 - `cues`: Character vector of cue dimensions to plot (1, 2, or 3 cues; defaults to as many cues as possible up to 3, selecting the first cues in the model).
 - `categories`: Optional character vector specifying a subset of category labels to evaluate and plot (defaults to the first category).
 - `decision_rule`: Decision rule label (defaults to `"proportional"`):
   - `"proportional"`: Response probability equals posterior probability $P(c \mid \mathbf{x})$ (probability matching).
   - `"criterion"`: Deterministic maximum a posteriori rule (winner-take-all classification boundary).
   - `"sampling"`: Stochastic sampling of category choices from posterior beliefs.
 - `aes`: Aesthetic layer type (`"contour"` for decision boundary contours, `"fill"` or `"fill-gradient"` for continuous response probability heatmaps, or `"fill-discrete"` for winner-take-all partition regions). Defaults to `"contour"`.
 - `limits`: Axis bounds, specified as a 2-element vector `c(min, max)` or named list `list(VOT = c(-30, 90), f0_semitones = c(-4, 6))`.
 - `resolution`: Grid density along each cue dimension (default `100`).
 
 Note: When all category priors are identical, the plot summary concisely indicates `all = {1 / K}`.

## 1D Categorization Functions {#1d-categorization-functions}

In 1D, categorization functions are classic psychometric identification curves:

```{r 1d_categorization_function, fig.width = 9.5, fig.height = 4}
model_1d <- example_mvg_ideal_observer(n_cues = 1)

# Left: Default proportional probability matching
p_cat_prop <- plot_categorization_function(
  model_1d, 
  categories = c("/b/", "/d/", "/g/")
) + labs(title = "Proportional Rule (Default)")

# Right: Criterion (deterministic argmax)
p_cat_crit <- plot_categorization_function(
  model_1d, 
  categories = c("/b/", "/d/", "/g/"),
  decision_rule = "criterion"
) + labs(title = "Criterion Rule (Step Boundaries)")

p_cat_prop + p_cat_crit + plot_layout(guides = "collect")
```

## 2D Categorization Functions {#2d-categorization-functions}

In 2D space, categorization functions partition the continuum into perceptual decision regions, separated by boundary contours (e.g. $P(c \mid \mathbf{x}) = 0.50$):

```{r 2d_categorization_function, fig.width = 9.5, fig.height = 4}
model_2d <- example_mvg_ideal_observer(n_cues = 2)

# Mode 1: Decision boundary contours for /b/ and /p/ (default aes = 'contour')
p_cat_cont <- plot_categorization_function(
  model_2d, 
  categories = c("/b/", "/p/"),
  limits = list(VOT = c(-30, 90), f0_semitones = c(-4, 6))
) + labs(title = "Decision Contours (default aes = 'contour')")

# Mode 2: Shaded posterior probability surface + contours for /b/ and /p/
p_cat_fill <- plot_categorization_function(
  model_2d,
  categories = c("/b/", "/p/"),
  aes = c("fill-gradient", "contour"),
  limits = list(VOT = c(-30, 90), f0_semitones = c(-4, 6))
) + labs(title = "Posterior Probability Landscape")

p_cat_cont + p_cat_fill
```

## Interactive Categorization Functions over 2D cues: 3D Categorization Surfaces {#interactive-3d-categorization-surfaces}

Just like 2D category densities can be rendered as 3D interactive elevation surfaces, 2D categorization functions can be rendered in 3D WebGL with `interactive = TRUE`. By default, interactive 3D surfaces use discrete fill (`aes = "fill-discrete"`), providing an intuitive visualization of the response probability landscape across the cue space. The $z$-axis represents the predicted response probability $P_{\text{resp}}(c \mid \mathbf{x}) \in [0, 1]$:

```{r 2d_interactive_categorization, fig.width = 8, fig.height = 6}
# 3D interactive response probability surface for category /b/
plot_categorization_function(
  model_2d,
  categories = "/b/",
  interactive = TRUE
)
```

## 3D Sliced Categorization Functions {#3d-sliced-categorization-functions}

For 3-cue models, `plot_categorization_function()` computes 2D decision contours and posterior response probabilities across slices of the third dimension, illustrating how boundaries shift as a function of the third acoustic cue:

```{r 3d_sliced_categorization, fig.width = 9.5, fig.height = 4}
# Multi-category decision contours across slices of F3
plot_categorization_function(
  m3,
  categories = c("b", "d", "g"),
  slices = c(2200, 2600, 3000),
  aes = c("fill-gradient", "contour")
)
```

---

# 3. Dimensionality Reduction & Marginal Projections {#dimensionality-reduction-and-marginal-projections}

When working with multidimensional models ($D \ge 3$), visualizing all cue dimensions simultaneously becomes challenging. `plot_categories()` and `plot_categorization_function()` support plotting 1D, 2D, or 3D subspaces by marginalizing over all omitted cues.

## Mathematical Principle of Marginalization {#mathematical-principle-of-marginalization}

When a user requests a 2D plot for a 3-cue model (for example, specifying `cues = c("F1", "F2")` for a model defined on `c("F1", "F2", "F3")`), **MVBeliefUpdatr** does **not** take an arbitrary slice at zero. Instead, it analytically **marginalizes out** the omitted dimensions by integrating over $\mathbb{R}$:
$$p(x_1, x_2 \mid c) = \int_{-\infty}^{\infty} p(x_1, x_2, x_3 \mid c) \, dx_3$$

For a multivariate Gaussian category representation with mean vector $\boldsymbol{\mu} \in \mathbb{R}^D$ and covariance matrix $\boldsymbol{\Sigma} \in \mathbb{R}^{D \times D}$, marginalizing out cue $k$ yields an exact Gaussian distribution with:
$$\boldsymbol{\mu}_{\text{proj}} = \boldsymbol{\mu}_I, \quad \boldsymbol{\Sigma}_{\text{proj}} = \boldsymbol{\Sigma}_{I, I}$$
where $I \subset \{1, \dots, D\}$ indexes the retained cue subset. This guarantees that all variances, covariances, and central density intervals displayed in the 2D plot reflect the true marginal distributions of the underlying model without distortion.

Similarly, for categorization functions, marginalizing out omitted cues corresponds to computing the posterior response probabilities of the marginalized category representations:
$$P(c \mid \mathbf{x}_I) = \frac{p(\mathbf{x}_I \mid c) P(c)}{\sum_{k} p(\mathbf{x}_I \mid k) P(k)}$$
reflecting the decision boundaries of an observer who only has access to the subset of cues $\mathbf{x}_I$, and thus has to marginalize over all other cues to make a categorization decision.

## Pairwise 2D Projections of a 3D Model {#pairwise-2d-projections-of-a-3d-model}

We can visualize all three pairwise combinations of a 3-cue phonetic space ($(F_1, F_2)$, $(F_1, F_3)$, and $(F_2, F_3)$) for both category densities and categorization functions:

### Pairwise Category Densities

```{r 3d_marginal_projections, fig.width = 11, fig.height = 3.8}
# 1. Height vs. Backness (F1 x F2)
p_f1_f2 <- plot_categories(m3, cues = c("F1", "F2"), aes = c("fill-gradient", "contour")) +
  labs(title = "Category likelihood", subtitle = "Marginalized over F3")

# 2. Height vs. Rhoticity (F1 x F3)
p_f1_f3 <- plot_categories(m3, cues = c("F1", "F3"), aes = c("fill-gradient", "contour")) +
  labs(title = "", subtitle = "Marginalized over F2")

# 3. Backness vs. Rhoticity (F2 x F3)
p_f2_f3 <- plot_categories(m3, cues = c("F2", "F3"), aes = c("fill-gradient", "contour")) +
  labs(title = "", subtitle = "Marginalized over F1")

# Assemble via patchwork
p_f1_f2 + p_f1_f3 + p_f2_f3 + plot_layout(guides = "collect")
```

### Pairwise Categorization Functions

```{r 3d_marginal_categorization, fig.width = 11, fig.height = 3.8}
# 1. F1 x F2 categorization
p_cat_f1_f2 <- plot_categorization_function(
  m3, 
  cues = c("F1", "F2"), 
  categories = c("b", "d", "g"), 
  aes = c("fill-gradient", "contour")
) + labs(title = "Categorization functions", subtitle = "Marginalized over F3")

# 2. F1 x F3 categorization
p_cat_f1_f3 <- plot_categorization_function(
  m3, 
  cues = c("F1", "F3"), 
  categories = c("b", "d", "g"), 
  aes = c("fill-gradient", "contour")
) + labs(title = "", subtitle = "Marginalized over F2")

# 3. F2 x F3 categorization
p_cat_f2_f3 <- plot_categorization_function(
  m3, 
  cues = c("F2", "F3"), 
  categories = c("b", "d", "g"), 
  aes = c("fill-gradient", "contour")
) + labs(title = "", subtitle = "Marginalized over F1")

# Assemble via patchwork
p_cat_f1_f2 + p_cat_f1_f3 + p_cat_f2_f3 + plot_layout(guides = "collect")
```

---

# 4. Visualizing Bayesian Belief Updating Models (`MVBU_Stanfit`) {#visualizing-bayesian-belief-updating-models}

Fitted **MVBeliefUpdatr** Stan models (`MVBU_Stanfit`) contain posterior MCMC draws reflecting uncertainty over category parameters across prior and exposure groups.

### Posterior Category Beliefs: Mean vs. Draw Uncertainty {#posterior-category-beliefs-mean-vs-draw-uncertainty}

Using `plot_categories()` on an `MVBU_Stanfit` object allows comparing point estimates against parameter uncertainty:

 - `sample`: Logical (`FALSE` by default to plot only posterior mean distributions, `TRUE` to overlay individual posterior MCMC draws).
 - `ndraws`: Integer specifying the number of posterior draws to sample and overlay with alpha transparency (default `50`).

```{r stanfit_load_and_plot_categories, fig.width = 9.5, fig.height = 4}
# Load pre-fitted 2D NIW ideal adaptor model
fit_path <- system.file("models/minimal-niw_ideal_adaptor-NIW-2cue.rds", package = "MVBeliefUpdatr")
if (nzchar(fit_path) && file.exists(fit_path)) {
  fit <- readRDS(fit_path)
} else if (file.exists("../tests/testthat/models/minimal-niw_ideal_adaptor-NIW-2cue.rds")) {
  fit <- readRDS("../tests/testthat/models/minimal-niw_ideal_adaptor-NIW-2cue.rds")
} else {
  fit <- readRDS("tests/testthat/models/minimal-niw_ideal_adaptor-NIW-2cue.rds")
}

# 1. Posterior Mean Distribution only (sample defaults to FALSE)
p_fit_mean <- plot_categories(fit) + labs(title = "Posterior Mean Beliefs")

# 2. Posterior Uncertainty Draws overlaid (sample = TRUE, ndraws = 30)
p_fit_sample <- plot_categories(fit, sample = TRUE, ndraws = 30) + labs(title = "Posterior Uncertainty Draws")

p_fit_mean + p_fit_sample + plot_layout(guides = "collect")
```

### Fitted 1D Categorization Function with Uncertainty {#fitted-1d-categorization-function-with-uncertainty}

`plot_categorization_function()` evaluates posterior response probability curves across exposure groups, with 95% posterior credible interval ribbons:

```{r stanfit_1d_categorization, fig.width = 8, fig.height = 3.5}
plot_categorization_function(
  fit, 
  cues = "VOT"
)
```

### Fitted 2D Categorization Function with Uncertainty {#fitted-2d-categorization-function-with-uncertainty}

`plot_categorization_function()` evaluates posterior response probability surfaces for the fitted exposure groups, with optional MCMC draw overlay:

```{r stanfit_categorization, fig.width = 7, fig.height = 5}
plot_categorization_function(
  fit, 
  cues = c("VOT", "f0_semitones"), 
  aes = c("fill-gradient", "contour"), 
  ndraws = 30
)
```

### Posterior Parameter Distributions (`plot_parameters`) {#posterior-parameter-distributions}

`plot_parameters()` extracts marginal posterior distributions for all model parameters:

 - `parameters`: Character vector of parameter names to plot (defaults to all parameters).
 - `categories`: Subset of categories to include (defaults to all categories).
 - `combine_into_single_plot`: Logical (defaults to `TRUE`), combining parameter subplots into a single unified patchwork layout. If `FALSE`, returns a named list of individual `ggplot` objects for custom arrangement.

```{r stanfit_parameters, fig.width = 8.5, fig.height = 5}
# Combined parameter distribution plot (combine_into_single_plot defaults to TRUE)
plot_parameters(fit)
```

### Parameter Correlation Heatmap (`plot_parameter_correlations`) {#parameter-correlation-heatmap}

`plot_parameter_correlations()` computes Pearson correlation coefficients between all parameter pairs across MCMC draws. Gray bounding boxes along the diagonal group parameters belonging to the same category:

 - `parameters`: Vector specifying a subset of parameter types to include in the correlation matrix (defaults to all parameters).
 - `groups`: Subset of experimental exposure/prior groups (defaults to `"prior"`).

```{r stanfit_correlations, fig.width = 7, fig.height = 6}
# Correlation heatmap for prior group parameters (groups defaults to "prior")
plot_parameter_correlations(fit)
```

### Pairwise Posterior Relationships (`plot_parameters_pairwise`) {#pairwise-posterior-relationships}

`plot_parameters_pairwise()` provides comprehensive MCMC diagnostics across parameter pairs:

 - **Lower triangle:** Scatterplots of posterior draws with LOESS trendlines
 - **Diagonal:** Marginal univariate posterior auto-densities
 - **Upper triangle:** 2D bivariate density contours (`geom_density_2d`)
 - `parameters`: Allows focusing the scatter matrix on specific parameters of interest (defaults to all parameters).
 - `groups`: Subset of experimental exposure/prior groups (defaults to `"prior"`).

```{r stanfit_pairwise, fig.width = 8, fig.height = 7}
# Pairwise posterior parameter matrix (groups defaults to "prior")
plot_parameters_pairwise(fit)
```

---

# 5. Parallelization and Performance Optimization {#parallelization-and-performance-optimization}

Evaluating continuous probability densities and response probability surfaces over fine 2D and 3D grids can be computationally demanding, especially for high-dimensional models or non-parametric exemplar models. **MVBeliefUpdatr** includes native multi-core parallelization across core evaluation functions to accelerate rendering.

### Supported Functions
The `parallel` and `n_cores` arguments are supported across the core workflow:

 - `plot_categories()`: Multi-category continuous grid evaluations and 3D multi-slice representations.
 - `plot_categorization_function()`: High-resolution grid evaluations of posterior response probabilities and decision landscapes.
 - `sample_observations()`: Parallel simulation of observations across categories.
 - Model fitting and MCMC sampling (`fit_stan_model()`, `update()`): Multi-chain parallel sampling via Stan.

### Requesting Parallel Execution
To request parallel evaluation, specify `parallel = TRUE`:

```r
# Parallel grid evaluation for high-resolution 3D sliced category plots
plot_categories(
  m3,
  cues = c("F1", "F2", "F3"),
  aes = c("fill-gradient", "contour"),
  resolution = 120,
  parallel = TRUE,
  n_cores = 4
)
```

When `n_cores` is omitted, the package automatically defaults to `max(1L, parallel::detectCores() - 1L)` (or the value configured via `options(mc.cores)`).

### Parallel Architecture & Execution Details
Under the hood:

 - On Unix-based systems (Linux and macOS), parallelization leverages forked process memory via `parallel::mclapply()`, which achieves near-zero inter-process communication (IPC) serialization overhead.
 - On Windows, a background socket cluster (`parallel::makeCluster()`) is utilized via `parallel::parLapply()`.
 - Evaluation grids are partitioned into contiguous row slices, evaluated independently across worker cores, and aggregated into unified tidy data frames prior to rendering.

---

# 6. Summary {#summary}

The plotting functions in **MVBeliefUpdatr** are intended to provide a consistent, modular interface for visualizing cognitive models at every level of complexity:

 - Use `plot_categories()` to inspect generative category structures, likelihood densities, and covariance geometry in 1D, 2D, and 3D.
 - Use `plot_categorization_function()` to explore posterior response probability landscapes and decision boundaries.
 - Leverage analytic marginalization to project high-dimensional models onto intuitive 2D cue pairs.
 - For fitted Stan models, combine `sample = TRUE`, `plot_parameters()`, `plot_parameter_correlations()`, and `plot_parameters_pairwise()` to evaluate parameter uncertainty and MCMC geometry.
 - Accelerate intensive grid evaluations seamlessly using `parallel = TRUE`.
